---
title: "English Premier League Monte Carlo Analysis"
author: "Megan Dunnahoo, Jasmine DeMeyer, Macey Dodd"
date: "12/16/2021"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
set.seed(400)
```


```{r data, include=FALSE}
S11_12 <- read.csv("2011-2012_Match.csv")
S12_13 <- read.csv("2012-2013_Match.csv")
S13_14 <- read.csv("2013-2014_Match.csv")
S14_15 <- read.csv("2014-2015_Match.csv")
```

Choose Chelsea as home team and Liverpool as away team (match date of )

```{r data_manip, include=FALSE}

# Keep only necessary columns
S11_12 <- S11_12[,2:7]
S12_13 <- S12_13[,2:7]
S13_14 <- S13_14[,2:7]
S14_15 <- S14_15[,2:7]

# Add id to keep order
S11_12$Id <- seq(1:nrow(S11_12))
S12_13$Id <- seq(381,760)
S13_14$Id <- seq(761,1140)
S14_15$Id <- seq(1141,1520)

Away_Function <- function(team_name) {
  
  # Subset and combine away team matches
  away <- S11_12[S11_12$AwayTeam == team_name,]
  away <- rbind(away, S12_13[S12_13$AwayTeam == team_name,])
  away <- rbind(away, S13_14[S13_14$AwayTeam == team_name,])
  away <- rbind(away, S14_15[S14_15$AwayTeam == team_name,])
  
  # Create new columns
  away$num_points <- rep(NA, length(away$FTR))
  away$outcome <- rep(NA, length(away$FTR))
  
  # Fill new columns
  for (i in 1:length(away$num_points)) {
    if (away$FTR[i] == "A") {
      away$num_points[i] <- 3
      away$outcome[i] <- "W"
    }
    if (away$FTR[i] == "H") {
      away$num_points[i] <- 0
      away$outcome[i] <- "L"
    }
    if (away$FTR[i] == "D") {
      away$num_points[i] <- 1
      away$outcome[i] <- "D"
    }
  }
  return(away)
}

Home_Function <- function(team_name) {
  
  # Subset and combine home team matches
  home <- S11_12[S11_12$HomeTeam == team_name,]
  home <- rbind(home, S12_13[S12_13$HomeTeam == team_name,])
  home <- rbind(home, S13_14[S13_14$HomeTeam == team_name,])
  home <- rbind(home, S14_15[S14_15$HomeTeam == team_name,])
  
  # Create new columns
  home$num_points <- rep(NA, length(home$FTR))
  home$outcome <- rep(NA, length(home$FTR))
  
  # Fill new columns
  for (i in 1:length(home$num_points)) {
    if (home$FTR[i] == "A") {
      home$num_points[i] <- 0
      home$outcome[i] <- "L"
    }
    if (home$FTR[i] == "H") {
      home$num_points[i] <- 3
      home$outcome[i] <- "W"
    }
    if (home$FTR[i] == "D") {
      home$num_points[i] <- 1
      home$outcome[i] <- "D"
    }
  }
  return(home)
}

teams <- unique(S11_12$HomeTeam)
for(i in teams) {
  nam_h <- paste(i, "Home", sep = "_")
  assign(nam_h, Home_Function(i))
  nam_a <- paste(i, "Away", sep = "_")
  assign(nam_a, Away_Function(i))
}

```


```{r, include=FALSE}
#Jasmine's test section

avg_goalsh <- sum(tail(Chelsea_Home$FTHG, 19))/19
avg_goals_concededh <- sum(tail(Chelsea_Home$FTAG, 19))/19

avg_goalsa <- sum(tail(Liverpool_Away$FTAG, 19))/19
avg_goals_concededa <- sum(tail(Liverpool_Away$FTHG, 19))/19

lam_home <- (avg_goalsh + avg_goals_concededa)/2 
lam_away <- (avg_goalsa + avg_goals_concededh)/2

away_df <- cbind(tail(Arsenal_Away, 19), tail(`Aston Villa_Away`,19), tail(Blackburn_Away, 19), tail(Bolton_Away, 19), tail(Chelsea_Away, 19), tail(Everton_Away, 19), tail(Fulham_Away, 19), tail(Liverpool_Away, 19), tail(`Man City_Away`, 19), tail(`Man United_Away`, 19), tail(Newcastle_Away, 19), tail(Norwich_Away, 19), tail(QPR_Away, 19), tail(Stoke_Away, 19), tail(Sunderland_Away, 19), tail(Swansea_Away, 19), tail(Tottenham_Away, 19), tail(`West Brom_Away`, 19), tail(Wigan_Away, 19), tail(Wolves_Away, 19))

home_df <- cbind(tail(Arsenal_Home, 19), tail(`Aston Villa_Home`,19), tail(Blackburn_Home, 19), tail(Bolton_Home, 19), tail(Chelsea_Home, 19), tail(Everton_Home, 19), tail(Fulham_Home, 19), tail(Liverpool_Home, 19), tail(`Man City_Away`, 19), tail(`Man United_Away`, 19), tail(Newcastle_Away, 19), tail(Norwich_Home, 19), tail(QPR_Home, 19), tail(Stoke_Home, 19), tail(Sunderland_Home, 19), tail(Swansea_Home, 19), tail(Tottenham_Home, 19), tail(`West Brom_Home`, 19), tail(Wigan_Home, 19), tail(Wolves_Home, 19))

```

```{r mc, include=FALSE}
away_df2 <- rbind(Arsenal_Away,`Aston Villa_Away`,Blackburn_Away,Bolton_Away,Chelsea_Away,Everton_Away,Fulham_Away, Liverpool_Away,`Man City_Away`,`Man United_Away`,Newcastle_Away,Norwich_Away,QPR_Away,Stoke_Away,Sunderland_Away, Swansea_Away,Tottenham_Away,`West Brom_Away`,Wigan_Away,Wolves_Away)

away_df2 %>% group_by(AwayTeam)

home_df2 <- rbind(Arsenal_Home,`Aston Villa_Home`,Blackburn_Home,Bolton_Home,Chelsea_Home,Everton_Home,Fulham_Home, Liverpool_Home,`Man City_Home`,`Man United_Home`,Newcastle_Home,Norwich_Home,QPR_Home,Stoke_Home,Sunderland_Home, Swansea_Home,Tottenham_Home,`West Brom_Home`,Wigan_Home,Wolves_Home)

home_df2 %>% group_by(HomeTeam)


mc_funct <- function(match_id, away_team, home_team) {
  home_df_tail <- home_df2 %>% filter(Id < match_id) %>% group_by(HomeTeam) %>% do(tail(., 19))
  away_df_tail <- away_df2 %>% filter(Id < match_id) %>% group_by(AwayTeam) %>% do(tail(., 19))
  
  home_t <- home_df_tail[home_df_tail$HomeTeam==home_team,]
  avg_goalsh <- sum(home_t$FTHG, 19)/19
  avg_goals_concededh <- sum(home_t$FTAG, 19)/19
  
  away_t <- away_df_tail[away_df_tail$AwayTeam==away_team,]
  avg_goalsa <- sum(away_t$FTAG, 19)/19
  avg_goals_concededa <- sum(away_t$FTHG, 19)/19
  
  lam_home_val <- (avg_goalsh + avg_goals_concededa)/2 
  lam_away_val <- (avg_goalsa + avg_goals_concededh)/2
  
  m <- 100000
  matches <- data.frame(Home=rep(NA,m), Away=rep(NA,m), ResultH=rep(NA,m), ResultA=rep(NA,m))
  for (i in 1:m) {
    home_goals <- rpois(1,lam_home_val)
    away_goals <- rpois(1,lam_away_val)
    matches[i,1] <- home_goals
    matches[i,2] <- away_goals
    if (home_goals > away_goals) {
      matches[i,3] <- "Win"
      matches[i,4] <- "Loss"
    }
    if (home_goals < away_goals) {
      matches[i,3] <- "Loss"
      matches[i,4] <- "Win"
    }
    if (home_goals == away_goals) {
      matches[i,3] <- "Draw"
      matches[i,4] <- "Draw"
    }
  }
  matches[,3] <- as.factor(matches[,3])
  matches[,4] <- as.factor(matches[,4])
  return(matches)
}

```

# MC Estimation

We estimated the lambda home and lambda away values, which are the expected goals scored by the home and away team respectively, using Monte Carlo. These lambda values are assumed to be independent Poisson distributed variables and are calculated using the last 19 matches for each team.     

$\lambda_{home}=\frac{\text{Average goals scored by home team + Average goals conceded by away team}}{2}$   
$\lambda_{away}=\frac{\text{Average goals scored by away team + Average goals conceded by home team}}{2}$

# Exploratory Plot of Outcomes Before Match (Chelsea vs Liverpool)

```{r graph1, echo=FALSE}
chel <- Home_Function("Chelsea")
chel$Date <- as.Date(chel$Date, "%d/%m/%y")
chel <- chel[1:74,]

liv <- Away_Function("Liverpool")
liv$Date <- as.Date(liv$Date, "%d/%m/%y")
liv <- liv[1:74,]

plot1 <- ggplot(chel) +
  geom_point(aes(Date, outcome, color=outcome)) + ggtitle("Chelsea")
plot2 <- ggplot(liv) +
  geom_point(aes(Date, outcome, color=outcome)) + ggtitle("Liverpool")
plot_grid(plot1, plot2, labels="AUTO", ncol=1)
```

# Prediction for Match 2014-2015 Chelsea (home) vs Liverpool (away)
Predictions are in terms of the home team   
Actual result was a draw

```{r pred, echo=FALSE}
# Takes a couple minutes to run
Match1497 <- mc_funct(match_id=1497, home_team="Chelsea", away_team="Liverpool")
res1497 <- table(t(Match1497$ResultH))
res1497 <- as.data.frame(res1497)
res1497$Percent <- res1497$Freq/100000 * 100
order <- c("Win", "Loss", "Draw")
res1497 <- res1497 %>%
  slice(match(Var1, order))
colnames(res1497) <- c("Match Result", "Occurances", "Percent")
res1497
```

# Exploratory Plot of Outcomes Before Match (Man City vs Newcastle)

```{r graph2, echo=FALSE}
man <- Home_Function("Man City")
man$Date <- as.Date(man$Date, "%d/%m/%y")
man <- man[1:38,]

new <- Away_Function("Newcastle")
new$Date <- as.Date(new$Date, "%d/%m/%y")
new <- new[1:38,]

plot3 <- ggplot(man) +
  geom_point(aes(Date, outcome, color=outcome)) + ggtitle("Manchester City")
plot4 <- ggplot(new) +
  geom_point(aes(Date, outcome, color=outcome)) + ggtitle("Newcastle")
plot_grid(plot3, plot4, labels="AUTO", ncol=1)
```

# Prediction for Match 2013-2014 Manchester City (home) vs Newcastle (away)
Predictions are in terms of the home team   
Actual result was a home win

```{r pred2, echo=FALSE}
# Takes a couple minutes to run
Match770 <- mc_funct(match_id=770, home_team="Man City", away_team="Newcastle")
res770 <- table(t(Match770$ResultH))
res770 <- as.data.frame(res770)
res770$Percent <- res770$Freq/100000 * 100
order <- c("Win", "Loss", "Draw")
res770 <- res770 %>%
  slice(match(Var1, order))
colnames(res770) <- c("Match Result", "Occurances", "Percent")
res770

```

