---
title: "Stat 400 Project - Group 4"
author: "Megan Dunnahoo, Jasmine DeMeyer, Macey Dodd"
output: beamer_presentation
---

```{r setup, include=FALSE}
library(tidyverse)
set.seed(400)
```

# Article 

We chose the article "Using Monte Carlo Simulation to Calculate Match Importance:
The Case of English Premier League" by Jiri Lahvicka. This article describes the process of using Monte Carlo simulations to predict the outcome of a match given the results of previous matches. It then goes further and uses Monte Carlo simulations to predict the final ranking of the teams in the English Premier League at the end of a season. 


```{r data, echo=FALSE}
S11_12 <- read.csv("2011-2012_Match.csv")
S12_13 <- read.csv("2012-2013_Match.csv")
S13_14 <- read.csv("2013-2014_Match.csv")
S14_15 <- read.csv("2014-2015_Match.csv")
```

# Soccer Information

What is necessary information on soccer? There are 20 teams in the English Premier League. Each team plays 19 away games and 19 home games?

# Data Manipulation

We wanted to use the specific variables FTR, FTAG, and FTHG, along with the identifier variables of Date, Away Team, and Home Team. There were no NA values in any of our selected variables.   

```{r data_manip, echo=FALSE, include=FALSE}

# Keep only necessary columns
S11_12 <- S11_12[,2:7]
S12_13 <- S12_13[,2:7]
S13_14 <- S13_14[,2:7]
S14_15 <- S14_15[,2:7]

# Add id to keep order
S11_12$Id <- seq(1:nrow(S11_12))
S12_13$Id <- seq(381,760)
S13_14$Id <- seq(761,1140)
S14_15$Id <- seq(1141,1520)

Away_Function <- function(team_name) {
  
  # Subset and combine away team matches
  away <- S11_12[S11_12$AwayTeam == team_name,]
  away <- rbind(away, S12_13[S12_13$AwayTeam == team_name,])
  away <- rbind(away, S13_14[S13_14$AwayTeam == team_name,])
  away <- rbind(away, S14_15[S14_15$AwayTeam == team_name,])
  
  # Create new columns
  away$num_points <- rep(NA, length(away$FTR))
  away$outcome <- rep(NA, length(away$FTR))
  
  # Fill new columns
  for (i in 1:length(away$num_points)) {
    if (away$FTR[i] == "A") {
      away$num_points[i] <- 3
      away$outcome[i] <- "W"
    }
    if (away$FTR[i] == "H") {
      away$num_points[i] <- 0
      away$outcome[i] <- "L"
    }
    if (away$FTR[i] == "D") {
      away$num_points[i] <- 1
      away$outcome[i] <- "D"
    }
  }
  return(away)
}

Home_Function <- function(team_name) {
  
  # Subset and combine home team matches
  home <- S11_12[S11_12$HomeTeam == team_name,]
  home <- rbind(home, S12_13[S12_13$HomeTeam == team_name,])
  home <- rbind(home, S13_14[S13_14$HomeTeam == team_name,])
  home <- rbind(home, S14_15[S14_15$HomeTeam == team_name,])
  
  # Create new columns
  home$num_points <- rep(NA, length(home$FTR))
  home$outcome <- rep(NA, length(home$FTR))
  
  # Fill new columns
  for (i in 1:length(home$num_points)) {
    if (home$FTR[i] == "A") {
      home$num_points[i] <- 0
      home$outcome[i] <- "L"
    }
    if (home$FTR[i] == "H") {
      home$num_points[i] <- 3
      home$outcome[i] <- "W"
    }
    if (home$FTR[i] == "D") {
      home$num_points[i] <- 1
      home$outcome[i] <- "D"
    }
  }
  return(home)
}

teams <- unique(S11_12$HomeTeam)
for(i in teams) {
  nam_h <- paste(i, "Home", sep = "_")
  assign(nam_h, Home_Function(i))
  nam_a <- paste(i, "Away", sep = "_")
  assign(nam_a, Away_Function(i))
}

```

```{r mc, echo=FALSE, include=FALSE}
away_df2 <- rbind(Arsenal_Away,`Aston Villa_Away`,Blackburn_Away,Bolton_Away,Chelsea_Away,Everton_Away,Fulham_Away, Liverpool_Away,`Man City_Away`,`Man United_Away`,Newcastle_Away,Norwich_Away,QPR_Away,Stoke_Away,Sunderland_Away, Swansea_Away,Tottenham_Away,`West Brom_Away`,Wigan_Away,Wolves_Away)

away_df2 %>% group_by(AwayTeam)

home_df2 <- rbind(Arsenal_Home,`Aston Villa_Home`,Blackburn_Home,Bolton_Home,Chelsea_Home,Everton_Home,Fulham_Home, Liverpool_Home,`Man City_Home`,`Man United_Home`,Newcastle_Home,Norwich_Home,QPR_Home,Stoke_Home,Sunderland_Home, Swansea_Home,Tottenham_Home,`West Brom_Home`,Wigan_Home,Wolves_Home)

home_df2 %>% group_by(HomeTeam)


mc_funct <- function(match_id, away_team, home_team) {
  home_df_tail <- home_df2 %>% filter(Id < match_id) %>% group_by(HomeTeam) %>% do(tail(., 19))
  away_df_tail <- away_df2 %>% filter(Id < match_id) %>% group_by(AwayTeam) %>% do(tail(., 19))
  
  home_t <- home_df_tail[home_df_tail$HomeTeam==home_team,]
  avg_goalsh <- sum(home_t$FTHG, 19)/19
  avg_goals_concededh <- sum(home_t$FTAG, 19)/19
  
  away_t <- away_df_tail[away_df_tail$AwayTeam==away_team,]
  avg_goalsa <- sum(away_t$FTAG, 19)/19
  avg_goals_concededa <- sum(away_t$FTHG, 19)/19
  
  lam_home_val <- (avg_goalsh + avg_goals_concededa)/2 
  lam_away_val <- (avg_goalsa + avg_goals_concededh)/2
  
  m <- 100000
  matches <- data.frame(Home=rep(NA,m), Away=rep(NA,m), ResultH=rep(NA,m), ResultA=rep(NA,m))
  for (i in 1:m) {
    home_goals <- rpois(1,lam_home_val)
    away_goals <- rpois(1,lam_away_val)
    matches[i,1] <- home_goals
    matches[i,2] <- away_goals
    if (home_goals > away_goals) {
      matches[i,3] <- "Win"
      matches[i,4] <- "Loss"
    }
    if (home_goals < away_goals) {
      matches[i,3] <- "Loss"
      matches[i,4] <- "Win"
    }
    if (home_goals == away_goals) {
      matches[i,3] <- "Draw"
      matches[i,4] <- "Draw"
    }
  }
  matches[,3] <- as.factor(matches[,3])
  matches[,4] <- as.factor(matches[,4])
  return(matches)
}

```

# MC Estimation

Estimating lambda values

lhome ¼ Average goals scored by home team þ Average goals conceded by away team
2
laway ¼ Average goals conceded by home team þ Average goals scored by away team
2

# Prediction for Match 2014-2015 Chelsea (home) vs Liverpool (away)
Predictions are in terms of the home team   
Actual result was a draw

```{r pred, echo=FALSE}
# Takes a couple minutes to run
Match1497 <- mc_funct(match_id=1497, home_team="Chelsea", away_team="Liverpool")
res1497 <- table(t(Match1497$ResultH))
res1497 <- as.data.frame(res1497)
res1497$Percent <- res1497$Freq/100000 * 100
order <- c("Win", "Loss", "Draw")
res1497 <- res1497 %>%
  slice(match(Var1, order))
colnames(res1497) <- c("Match Result", "Occurances", "Percent")
res1497

```

# Prediction for Match 2013-2014 Manchester City (home) vs Newcastle (away)
Predictions are in terms of the home team   
Actual result was a home win

```{r pred2, echo=FALSE}
# Takes a couple minutes to run
Match770 <- mc_funct(match_id=770, home_team="Man City", away_team="Newcastle")
res770 <- table(t(Match770$ResultH))
res770 <- as.data.frame(res770)
res770$Percent <- res770$Freq/100000 * 100
order <- c("Win", "Loss", "Draw")
res770 <- res770 %>%
  slice(match(Var1, order))
colnames(res770) <- c("Match Result", "Occurances", "Percent")
res770

```

# Going Forward 

In the next week we plan to continue to expand on our exploration of this data. We plan to integrate code into our Monte Carlo function to predict the final rank of any two teams we have chosen at the time of a specific match. We also hope to add to our final contingency table a variable that consists of the probability of the ranks for these two teams to add to the number of wins, losses, and draws they had in our simulations.
